\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-11 analysis/paper/paper.tex
edit third case study

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx}

@@

\begin{table}[ht]
\centering
\begin{tabular}{ccc}
  \hline
Variable & Description & Statistical summary \\ 
  \hline
Sex & Sex & M=Male (100) \\ 
   &  & F=Female (77) \\ 
  D & Disease & ALL (73) \\ 
   &  & AML (104) \\ 
  Phase & Phase & CR1 (47) \\ 
   &  & CR2 (45) \\ 
   &  & CR3 (12) \\ 
   &  & Relapse (73) \\ 
  Source & Type of transplant & BM+PB (21) \\ 
   &  & PB (156) \\ 
  Age & Age of patient (years) & 4–62 \\ 
   &  & 30.47 (13.04) \\ 
  Ftime & Failure time (months) & 0.13–131.77 \\ 
   &  & 20.28 (30.78) \\ 
  Status & Status indicator & 0=censored (46) \\ 
   &  & 1=relapse (56) \\ 
   &  & 2=competing event (75) \\ 
   \hline
\end{tabular}
\caption{Baseline characteristics of patients in the stem-cell transplant study.}
\label{tab:table1bmtcrr}
\end{table}

@@

\hypertarget{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}{%
\section{Case study 3--Study to Understand Prognoses Preferences
Outcomes and Risks of
Treatment}\label{case-study-3study-to-understand-prognoses-preferences-outcomes-and-risks-of-treatment}}

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset (REF from main site). the SUPPORT
dataset tracks death for individuals who are considered seriously ill
within a hospital. Imputation was conducted using the default methods in
the mice package in R, available on CRAN. Before imputation, there was a
total of 1000 individuals and 35 variables. After imputation, we have
474 observations where 67.5\% of individuals in the study experience the
event of interest. The SUPPORT dataset will be used to demonstrate
regularization within casebase (REF), while comparing their absolute
risk predictions for a new individual. A description of each variable
can be found in Table\textasciitilde{}\ref{tab:support1} and a breakdown
of each categorical variable in
Table\textasciitilde{}\ref{tab:support2}. The data was downloaded from
the main author's website, and it is also available as part of the
casebase package.

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset, taken from REF.}
\label{tab:support2}
\end{table}

Hospital death as a covariate was removed, as this is directly
informative of death. The glmnet package(REF) on CRAN was used to
introduce elastic net, lasso and ridge regressions to casebase. To
visualize the effect of imputation on incidence density, we can refer to
the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and
post-imputation population time plot in Figure \ref{fig:ptPostI}. The
incidence density is consistent between imputed and complete datasets.

@@

\begin{CodeInput}
R> #demonstrating incidence density before imputation
R> plot(casebase::popTime(as.data.frame(support),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> #demonstrating incidence density after imputation
R> plot(casebase::popTime(as.data.frame(workingCompleteData),time = "d.time",event = "death"))
\end{CodeInput}
\end{CodeChunk}

Our main objective is to compute the absolute risk of death for a given
set of covariates, using regularization when fitting our model. First,
we fit a smooth hazard to the data; for the sake of this example, we
opted for a linear term for time. As casebase makes use of the glmnet
package, we interact with the fitsmoothhazard.fit function using a
matrix interface, where y contains the time and event variables and x
contains all other variables we would like to include in the model. For
the SUPPORT dataset, all factor variables were converted into dummy
variables. Here, we used lasso penalization by setting alpha to 1.

@@

\begin{CodeInput}
R> #Using casebase's glmnet implementation, using splines and a case to base ratio of 100,
R> #to determine the hazard function
R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
R>                              formula_time = ~d.time,alpha=1,ratio=100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model and use the absoluteRisk function to integrate
and retrieve our absolute risk function.

@@

\begin{CodeInput}
R> #Estimating the absolute risk curve using the newData parameter.
R> casebaseAbsolute=casebase::absoluteRisk(cb.Model,time = seq(1,max(completeData$d.time), 0.5),newdata = t(newData) ,
R>                               s="lambda.1se",method=c("numerical"))
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of cox regression,
and a kaplan-meier survival curve. The regularized cox regression
absolute risk curve required two extra steps to retrieve. Coxnet from
the glmnet package has tools to fit a regularized cox model, but
requires that a survival object skeleton of the selected variables, so
that the survival package tools can be used to retrieve the absolute
risk. We handle this by fitting a regularized cox model with the glmnet
package, and a cox model from the survival package with the selected
variables from the regularized fit. This second model serves as a
skeleton that permits the use of survival package functions. The
coefficients from the regularized model will replace the coefficients in
our skeleton. the resulting model will have the correct coefficients
when integrating, but will have an incorrect standard error. For the
purposes of this case study, we are only interested in the absolute risk
curve itself and not the standard error.

@@

\begin{CodeInput}
R> #Create u and xCox, which will be used when fitting Cox with glmnet.
R> u=survival::Surv(time = as.numeric(y[,2]), event = as.factor(y[,1]))
R> xCox=as.matrix(sparse.model.matrix(death~ .-d.time,data=completeData))[-c(sam),]
R> #hazard for cox using glmnet
R> coxglmFit=glmnet::cv.glmnet(x=xCox,y=u, family="cox",alpha=1)
R> #convergence demonstrated in plot
R> plot(coxglmFit)
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
R> #creating a new dataset that only contains the covariates chosen through glmnet.
R> #cleanCoxData<- as.data.frame(cbind(as.numeric(workingData$d.time),as.factor(workingData$death), xCox[,nonzero_covariate_cox$X1]))
R> cleanCoxData<-as.data.frame(cbind(as.numeric(y[,2]),as.numeric(y[,1]),xCox[,nonzero_covariate_cox$X1]))
R> #newDataCox<-xCox[sam,nonzero_covariate_cox$X1]
R> #fitting a cox model using regular estimation, however we will not keep it.
R> #this is used more as an object place holder.
R> coxFit <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = cleanCoxData)
R> #The coefficients of this object will be replaced with the estimates from coxglmFit.
R> #Doing so makes it so that everything is invalid aside from the coefficients.
R> #In this case, all we need to estimate the absolute risk is the coefficients.
R> #Std. error would be incorrect here, if we were to draw error bars.
R> coxFit$coefficients<-nonzero_coef_cox@x
R> #Fitting absolute risk curve for cox+glmnet
R> newDataCox=newData[nonzero_covariate_cox$X1-1]
R> 
R> abCoxFit<-survival::survfit(coxFit,newdata=as.data.frame(t(newDataCox)),time = seq(0,max(completeData$d.time), 1),type="breslow")
\end{CodeInput}
\end{CodeChunk}

We used the survival package to calculate the kaplan-meier absolute risk
function.

@@

\begin{CodeInput}
R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
R> # (Kaplan-Meier risk curve)
R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
\end{CodeInput}
\end{CodeChunk}

Now that our three absolute risk functions have been calculated, we
compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
decrease the absolute risk by the end of the study, in comparison to
kaplan-meier.

@@

\begin{CodeInput}
R> # A plot to compare all three Absolute risk (Commulative Incidence)
R> plot(casebaseAbsolute,type='l',col="black",lwd=2,main="Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves",xlab="Survival-Time",ylab="Cumulative Incidence",ylim=c(0,1),xlim=)
R> lines(abCoxFit ,col="red",fun="event",lwd=3,conf.int = FALSE)
R> lines(abKm ,col="Blue",fun="event",lwd=3,conf.int = FALSE)
R> legend("bottom", 
R>        legend = c( "Casebase (Lasso+linear)","semi-parametric (Cox)+glmnet","KM curve"), 
R>        col = c("black","red","Blue"),
R>        lty = c(1, 1, 1), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

The purpose of the next case study is to show how \pkg{casebase} can
also be used for survival analysis problems that do not share the
properties above. To this end, we will use the Stanford Heart Transplant
Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
use the version available in the \pkg{survival} package.

@@

In the following table we provide a comparison between the Cox model and
case-base sampling:

@@

This report was generated on 2019-12-10 12:28:40 using the following
computational environment and dependencies:

@@

#> - Session info ----------------------------------------------------------

@@

#>  date     2019-12-10                  

@@

#> - Packages --------------------------------------------------------------
#>  package     * version    date       lib source                          
#>  assertthat    0.2.1      2019-03-21 [1] CRAN (R 3.6.1)                  
#>  backports     1.1.5      2019-10-02 [1] CRAN (R 3.6.1)                  
#>  broom         0.5.2      2019-04-07 [1] CRAN (R 3.6.1)                  
#>  callr         3.3.1      2019-07-18 [1] CRAN (R 3.6.1)                  
#>  casebase    * 0.2.1.9001 2019-07-29 [1] local                           
#>  cellranger    1.1.0      2016-07-27 [1] CRAN (R 3.5.0)                  
#>  cli           1.1.0      2019-03-19 [1] CRAN (R 3.6.1)                  
#>  codetools     0.2-16     2018-12-24 [4] CRAN (R 3.5.2)                  
#>  colorspace    1.4-1      2019-03-18 [1] CRAN (R 3.6.1)                  
#>  crayon        1.3.4      2017-09-16 [1] CRAN (R 3.5.0)                  
#>  data.table    1.12.2     2019-04-07 [1] CRAN (R 3.6.1)                  
#>  desc          1.2.0      2018-05-01 [1] CRAN (R 3.5.0)                  
#>  devtools      2.1.0      2019-07-06 [1] CRAN (R 3.6.1)                  
#>  digest        0.6.22     2019-10-21 [1] CRAN (R 3.6.1)                  
#>  dplyr       * 0.8.3      2019-07-04 [1] CRAN (R 3.6.1)                  
#>  ellipsis      0.3.0      2019-09-20 [1] CRAN (R 3.6.1)                  
#>  evaluate      0.14       2019-05-28 [1] CRAN (R 3.6.1)                  
#>  forcats     * 0.4.0      2019-02-17 [1] CRAN (R 3.6.1)                  
#>  foreach     * 1.4.4      2017-12-12 [1] CRAN (R 3.5.0)                  
#>  fs            1.3.1      2019-05-06 [1] CRAN (R 3.6.1)                  
#>  generics      0.0.2      2018-11-29 [1] CRAN (R 3.6.1)                  
#>  ggplot2     * 3.2.1      2019-08-10 [1] CRAN (R 3.6.1)                  
#>  glmnet      * 2.0-18     2019-05-20 [1] CRAN (R 3.6.1)                  
#>  glue          1.3.1      2019-03-12 [1] CRAN (R 3.6.1)                  
#>  gtable        0.3.0      2019-03-25 [1] CRAN (R 3.6.1)                  
#>  haven         2.1.1      2019-07-04 [1] CRAN (R 3.6.1)                  
#>  hms           0.5.2      2019-10-30 [1] CRAN (R 3.6.1)                  
#>  htmltools     0.3.6      2017-04-28 [1] CRAN (R 3.5.0)                  
#>  httr          1.4.1      2019-08-05 [1] CRAN (R 3.6.1)                  
#>  iterators     1.0.9      2017-12-12 [1] CRAN (R 3.5.0)                  
#>  jsonlite      1.6        2018-12-07 [1] CRAN (R 3.6.1)                  
#>  knitr         1.23       2019-05-18 [1] CRAN (R 3.6.1)                  
#>  labeling      0.3        2014-08-23 [1] CRAN (R 3.5.0)                  
#>  lattice       0.20-38    2018-11-04 [4] CRAN (R 3.5.1)                  
#>  lazyeval      0.2.2      2019-03-15 [1] CRAN (R 3.6.1)                  
#>  lifecycle     0.1.0      2019-08-01 [1] CRAN (R 3.6.1)                  
#>  lubridate   * 1.7.4      2018-04-11 [1] CRAN (R 3.5.0)                  
#>  magrittr    * 1.5        2014-11-22 [1] CRAN (R 3.5.0)                  
#>  Matrix      * 1.2-18     2019-11-27 [4] CRAN (R 3.6.1)                  
#>  memoise       1.1.0      2017-04-21 [1] CRAN (R 3.5.0)                  
#>  mgcv          1.8-24     2018-06-18 [1] CRAN (R 3.5.1)                  
#>  modelr        0.1.5      2019-08-08 [1] CRAN (R 3.6.1)                  
#>  munsell       0.5.0      2018-06-12 [1] CRAN (R 3.6.1)                  
#>  nlme          3.1-142    2019-11-07 [4] CRAN (R 3.6.1)                  
#>  pillar        1.4.2      2019-06-29 [1] CRAN (R 3.6.1)                  
#>  pkgbuild      1.0.3      2019-03-20 [1] CRAN (R 3.6.1)                  
#>  pkgconfig     2.0.3      2019-09-22 [1] CRAN (R 3.6.1)                  
#>  pkgload       1.0.2      2018-10-29 [1] CRAN (R 3.6.1)                  
#>  prettyunits   1.0.2      2015-07-13 [1] CRAN (R 3.5.2)                  
#>  processx      3.4.1      2019-07-18 [1] CRAN (R 3.6.1)                  
#>  ps            1.3.0      2018-12-21 [1] CRAN (R 3.6.1)                  
#>  purrr       * 0.3.3      2019-10-18 [1] CRAN (R 3.6.1)                  
#>  R6            2.4.1      2019-11-12 [1] CRAN (R 3.6.1)                  
#>  Rcpp          1.0.3      2019-11-08 [1] CRAN (R 3.6.1)                  
#>  readr       * 1.3.1      2018-12-21 [1] CRAN (R 3.6.1)                  
#>  readxl        1.3.1      2019-03-13 [1] CRAN (R 3.6.1)                  
#>  remotes       2.1.0      2019-06-24 [1] CRAN (R 3.6.1)                  
#>  rlang         0.4.1      2019-10-24 [1] CRAN (R 3.6.1)                  
#>  rmarkdown     1.16       2019-10-01 [1] CRAN (R 3.6.1)                  
#>  rprojroot     1.3-2      2018-01-03 [1] CRAN (R 3.5.0)                  
#>  rstudioapi    0.10       2019-03-19 [1] CRAN (R 3.6.1)                  
#>  rticles       0.9.1      2019-07-22 [1] Github (rstudio/rticles@8d56fc6)
#>  rvest         0.3.5      2019-11-08 [1] CRAN (R 3.6.1)                  
#>  scales        1.0.0      2018-08-09 [1] CRAN (R 3.6.1)                  
#>  sessioninfo   1.1.1      2018-11-05 [1] CRAN (R 3.6.1)                  
#>  stringi       1.4.3      2019-03-12 [1] CRAN (R 3.6.1)                  
#>  stringr     * 1.4.0      2019-02-10 [1] CRAN (R 3.6.1)                  
#>  survival    * 2.44-1.1   2019-04-01 [4] CRAN (R 3.6.1)                  
#>  testthat      2.2.0      2019-07-22 [1] CRAN (R 3.6.1)                  
#>  tibble      * 2.1.3      2019-06-06 [1] CRAN (R 3.6.1)                  
#>  tidyr       * 1.0.0      2019-09-11 [1] CRAN (R 3.6.1)                  
#>  tidyselect    0.2.5      2018-10-11 [1] CRAN (R 3.5.2)                  
#>  tidyverse   * 1.2.1      2017-11-14 [1] CRAN (R 3.5.0)                  
#>  usethis       1.5.1      2019-07-04 [1] CRAN (R 3.6.1)                  
#>  vctrs         0.2.0      2019-07-05 [1] CRAN (R 3.6.1)                  
#>  VGAM          1.1-1      2019-02-18 [1] CRAN (R 3.6.1)                  
#>  withr         2.1.2      2018-03-15 [1] CRAN (R 3.5.0)                  
#>  xfun          0.8        2019-06-25 [1] CRAN (R 3.6.1)                  
#>  xml2          1.2.2      2019-08-09 [1] CRAN (R 3.6.1)                  
#>  yaml          2.2.0      2018-07-25 [1] CRAN (R 3.6.1)                  

>>>>

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \DeclareUnicodeCharacter{2500}{-}

@@

\begin{table}[ht]
\centering
\begin{tabular}{ccc}
  \hline
Variable & Description & Statistical summary \\ 
  \hline
Sex & Sex & M=Male (100) \\ 
   &  & F=Female (77) \\ 
  D & Disease & ALL (73) \\ 
   &  & AML (104) \\ 
  Phase & Phase & CR1 (47) \\ 
   &  & CR2 (45) \\ 
   &  & CR3 (12) \\ 
   &  & Relapse (73) \\ 
  Source & Type of transplant & BM+PB (21) \\ 
   &  & PB (156) \\ 
  Age & Age of patient (years) & 4--62 \\ 
   &  & 30.47 (13.04) \\ 
  Ftime & Failure time (months) & 0.13--131.77 \\ 
   &  & 20.28 (30.78) \\ 
  Status & Status indicator & 0=censored (46) \\ 
   &  & 1=relapse (56) \\ 
   &  & 2=competing event (75) \\ 
   \hline
\end{tabular}
\caption{Baseline characteristics of patients in the stem-cell transplant study.}
\label{tab:table1bmtcrr}
\end{table}

@@

\hypertarget{case-study-3support-data}{%
\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset \citeyearpar{knaus1995support}. The
SUPPORT dataset tracks death for individuals who are considered
seriously ill within a hospital. Imputation was conducted using the
default methods in the \pkg{mice} package in \proglang{R}, available on
CRAN.
\textcolor{red}{Why was this method used? The Vanderbilt University website lists describes another method for imputation that we could have used.}
Before imputation, there was a total of 1000 individuals and 35
variables. After imputation, we have 474 observations where 67.5\% of
individuals in the study experience the event of interest. The SUPPORT
dataset will be used to demonstrate regularization within casebase,
while comparing their absolute risk predictions for a new individual. A
description of each variable can be found in
Table\textasciitilde{}\ref{tab:support1} and a breakdown of each
categorical variable in Table\textasciitilde{}\ref{tab:support2}. The
data was downloaded from the main author's website, and it is also
available as part of the \pkg{casebase} package.

@@

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset.}
\label{tab:support2}
\end{table}

Hospital death as a covariate was removed, as this is directly
informative of death. The \pkg{glmnet} package on CRAN was used to
introduce elastic net, lasso and ridge regressions to \pkg{casebase}. To
visualize the effect of imputation on incidence density, we can refer to
the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and
post-imputation population time plot in Figure \ref{fig:ptPostI}. The
incidence density is consistent between imputed and complete datasets.

@@

\begin{CodeInput}
R> popTime_pre <- popTime(as.data.frame(support),
R>                        time = "d.time", event = "death")
R> plot(popTime_pre)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> popTime_post <- popTime(as.data.frame(workingCompleteData),
R>                         time = "d.time", event = "death")
R> plot(popTime_post)
\end{CodeInput}
\end{CodeChunk}

Our main objective is to estimate the cumulative incidence of death for
a given covariate profile. Given th number of covariates available, we
will do this using regularized estimation. First, we fit the hazard
function; for the sake of simplicity, we opted for a linear term for
time. As \pkg{casebase} makes use of the \pkg{glmnet} package, we will
use the function \code{fitsmoothhazard.fit}, which follows the syntax
from \code{glmnet}. Specifically, we input a matrix \code{y}, which
contains the time and event variables, and a matrix \code{x}, which
contains all other variables we would like to include in the model. For
the SUPPORT dataset, all factor variables were converted into dummy
variables. Here, we used lasso penalization by setting \code{alpha} to
1.

@@

\begin{CodeInput}
R> cb.Model <- fitSmoothHazard.fit(x, y, family = "glmnet", time = "d.time", 
R>                                 event = "death", formula_time = ~ d.time, 
R>                                 alpha = 1, ratio = 100)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model object and use the \code{absoluteRisk} function
to compute our estimate of the cumulative incidence function.

@@

\begin{CodeInput}
R> casebaseAbsolute <- absoluteRisk(cb.Model,
R>                                  time = seq(1, max(completeData$d.time), 0.5),
R>                                  newdata = t(newData),
R>                                  s = "lambda.1se", method = "numerical")
\end{CodeInput}
\end{CodeChunk}

We will compare this curve to a regularized version of Cox regression,
and a Kaplan-Meier survival curve. Using the \pkg{glmnet} package, we
can fit the Cox regression model using regularized estimation. However,
\pkg{glmnet} does not provide any functionality for estimating the
baseline hazard, and the functionalities in the \pkg{survival} package
do not work on the output from \code{cv.glmnet}. To circumvent this
issue and estimate the baseline hazard, we followed the following steps.
First, we fitted a non-regularized Cox model using only the variables
selected by \pkg{glmnet}; the baseline hazard will be estimated using
this object. Then, we replaced the estimates of the regression
coefficients by the estimates from \pkg{glmnet}. Finally, this modified
\code{coxph} object is fed to the \code{survival::survfit} function to
obtain the estimated survival function.

@@

\begin{CodeInput}
R> # Create u and xCox, which will be used when fitting Cox with glmnet.
R> u <- survival::Surv(time = as.numeric(y[,2]), 
R>                     event = as.factor(y[,1]))
R> xCox <- as.matrix(sparse.model.matrix(death~ . - d.time,
R>                                       data = completeData))[-c(sam),]
R> # hazard for cox using glmnet
R> coxglmFit <- glmnet::cv.glmnet(x = xCox, y = u,
R>                                family = "cox", alpha = 1)
R> # convergence demonstrated in plot
R> # plot(coxglmFit)
R> #taking the coefficient estimates for later use
R> nonzero_covariate_cox <- predict(coxglmFit, type = "nonzero", s = "lambda.1se")
R> nonzero_coef_cox <- coef(coxglmFit, s = "lambda.1se")
R> # creating a new dataset that only contains the covariates chosen through glmnet.
R> cleanCoxData <- as.data.frame(cbind(as.numeric(y[,2]),
R>                                     as.numeric(y[,1]),
R>                                     xCox[, nonzero_covariate_cox$X1]))
R> # fitting a cox model using regular estimation, however we will not keep it.
R> # this is used more as an object place holder.
R> coxFit <- survival::coxph(Surv(time = V1, event = V2) ~ ., 
R>                           data = cleanCoxData)
R> # The coefficients of this object will be replaced 
R> # with the estimates from coxglmFit.
R> coxFit$coefficients <- nonzero_coef_cox@x
R> # Fitting CI curve for cox+glmnet
R> newDataCox <- newData[nonzero_covariate_cox$X1 - 1]
R> 
R> abCoxFit <- survival::survfit(coxFit, newdata = as.data.frame(t(newDataCox)),
R>                               time = seq(0, max(completeData$d.time), 1),
R>                               type = "breslow")
\end{CodeInput}
\end{CodeChunk}

We used the \pkg{survival} package to calculate the Kaplan-Meier
estimate of the survival function.

@@

\begin{CodeInput}
R> km <- survival::Surv(time = completeData$d.time, 
R>                      event = completeData$death)
R> abKm <- survival::survfit(km ~ 1,
R>                           type = 'kaplan-meier')
\end{CodeInput}
\end{CodeChunk}

Now that our three cumulative incidence functions have been calculated,
we compare them all in Figure \ref{fig:abSC}. Both \pkg{glmnet} and
\pkg{casebase} have a lower estimated cumulative incidence by the end of
the study, when compared to Kaplan-Meier.

@@

\begin{CodeInput}
R> plot(casebaseAbsolute, type='l', col = "black", lwd = 2, 
R>      main = "Support- CaseBase vs. Cox+glmnet vs KM Absolute Risk Curves", 
R>      xlab = "Survival-Time", ylab = "Cumulative Incidence", ylim = c(0,1))
R> lines(abCoxFit, col = "red", fun = "event", lwd = 3, conf.int = FALSE)
R> lines(abKm, col = "blue", fun = "event", lwd = 3, conf.int = FALSE)
R> legend("bottom", 
R>        legend = c("casebase", "Cox+glmnet", "KM curve"), 
R>        col = c("black", "red", "blue"), lty = c(1, 1, 1), bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

The purpose of the next case study is to show how \pkg{casebase} can
also be used for survival analysis problems that do not share the
properties above. To this end, we will use the Stanford Heart Transplant
Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
use the version available in the \pkg{surivival} package.

@@

@@

This report was generated on 2019-12-11 10:31:00 using the following
computational environment and dependencies:

@@

#> ─ Session info ──────────────────────────────────────────────────────────

@@

#>  date     2019-12-11                  

@@

#> ─ Packages ──────────────────────────────────────────────────────────────
#>  package     * version    date       lib
#>  assertthat    0.2.1      2019-03-21 [1]
#>  backports     1.1.5      2019-10-02 [1]
#>  broom         0.5.2      2019-04-07 [1]
#>  callr         3.3.1      2019-07-18 [1]
#>  casebase    * 0.2.1.9001 2019-12-11 [1]
#>  cellranger    1.1.0      2016-07-27 [1]
#>  cli           2.0.0      2019-12-09 [1]
#>  codetools     0.2-16     2018-12-24 [4]
#>  colorspace    1.4-1      2019-03-18 [1]
#>  crayon        1.3.4      2017-09-16 [1]
#>  data.table    1.12.8     2019-12-09 [1]
#>  desc          1.2.0      2018-05-01 [1]
#>  devtools      2.1.0      2019-07-06 [1]
#>  digest        0.6.23     2019-11-23 [1]
#>  dplyr       * 0.8.3      2019-07-04 [1]
#>  ellipsis      0.3.0      2019-09-20 [1]
#>  evaluate      0.14       2019-05-28 [1]
#>  fansi         0.4.0      2018-10-05 [1]
#>  farver        2.0.1      2019-11-13 [1]
#>  forcats     * 0.4.0      2019-02-17 [1]
#>  foreach     * 1.4.4      2017-12-12 [1]
#>  fs            1.3.1      2019-05-06 [1]
#>  generics      0.0.2      2018-11-29 [1]
#>  ggplot2     * 3.2.1      2019-08-10 [1]
#>  glmnet      * 2.0-18     2019-05-20 [1]
#>  glue          1.3.1      2019-03-12 [1]
#>  gtable        0.3.0      2019-03-25 [1]
#>  haven         2.1.1      2019-07-04 [1]
#>  hms           0.5.2      2019-10-30 [1]
#>  htmltools     0.3.6      2017-04-28 [1]
#>  httr          1.4.1      2019-08-05 [1]
#>  iterators     1.0.9      2017-12-12 [1]
#>  jsonlite      1.6        2018-12-07 [1]
#>  knitr         1.23       2019-05-18 [1]
#>  labeling      0.3        2014-08-23 [1]
#>  lattice       0.20-38    2018-11-04 [4]
#>  lazyeval      0.2.2      2019-03-15 [1]
#>  lifecycle     0.1.0      2019-08-01 [1]
#>  lubridate   * 1.7.4      2018-04-11 [1]
#>  magrittr    * 1.5        2014-11-22 [1]
#>  Matrix      * 1.2-18     2019-11-27 [4]
#>  memoise       1.1.0      2017-04-21 [1]
#>  mgcv          1.8-24     2018-06-18 [1]
#>  modelr        0.1.5      2019-08-08 [1]
#>  munsell       0.5.0      2018-06-12 [1]
#>  nlme          3.1-142    2019-11-07 [4]
#>  pillar        1.4.2      2019-06-29 [1]
#>  pkgbuild      1.0.3      2019-03-20 [1]
#>  pkgconfig     2.0.3      2019-09-22 [1]
#>  pkgload       1.0.2      2018-10-29 [1]
#>  prettyunits   1.0.2      2015-07-13 [1]
#>  processx      3.4.1      2019-07-18 [1]
#>  ps            1.3.0      2018-12-21 [1]
#>  purrr       * 0.3.3      2019-10-18 [1]
#>  R6            2.4.1      2019-11-12 [1]
#>  Rcpp          1.0.3      2019-11-08 [1]
#>  readr       * 1.3.1      2018-12-21 [1]
#>  readxl        1.3.1      2019-03-13 [1]
#>  remotes       2.1.0      2019-06-24 [1]
#>  rlang         0.4.2      2019-11-23 [1]
#>  rmarkdown     1.16       2019-10-01 [1]
#>  rprojroot     1.3-2      2018-01-03 [1]
#>  rstudioapi    0.10       2019-03-19 [1]
#>  rticles       0.9.1      2019-07-22 [1]
#>  rvest         0.3.5      2019-11-08 [1]
#>  scales        1.1.0      2019-11-18 [1]
#>  sessioninfo   1.1.1      2018-11-05 [1]
#>  stringi       1.4.3      2019-03-12 [1]
#>  stringr     * 1.4.0      2019-02-10 [1]
#>  survival    * 2.44-1.1   2019-04-01 [4]
#>  testthat      2.2.0      2019-07-22 [1]
#>  tibble      * 2.1.3      2019-06-06 [1]
#>  tidyr       * 1.0.0      2019-09-11 [1]
#>  tidyselect    0.2.5      2018-10-11 [1]
#>  tidyverse   * 1.2.1      2017-11-14 [1]
#>  usethis       1.5.1      2019-07-04 [1]
#>  vctrs         0.2.0      2019-07-05 [1]
#>  VGAM          1.1-2      2019-11-21 [1]
#>  withr         2.1.2      2018-03-15 [1]
#>  xfun          0.8        2019-06-25 [1]
#>  xml2          1.2.2      2019-08-09 [1]
#>  yaml          2.2.0      2018-07-25 [1]
#>  zeallot       0.1.0      2018-01-28 [1]
#>  source                                  
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  Github (sahirbhatnagar/casebase@e2e73fe)
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.5.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.0)                          
#>  CRAN (R 3.5.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.5.2)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
#>  CRAN (R 3.6.1)                          
